Collective excitations in competing phases in two and three dimensions

### 2024-05-21 paper/main.tex
x
uni

We start with an operator set $\mathcal{B}$ that ideally is complete with respect to commutation with 
the Hamiltonian $H$, i.e., any commutator $[H, A]$ for any $A \in \mathcal{B}$ can be represented by linear combinations of operators in $\mathcal{B}$.
Then, we can express any time-dependent operator in this set by
\begin{equation}
    \label{eqn:time_dependent_operator}
    A(t) = \sum_n a_n(t) A_n,
\end{equation}
where the coefficients $c_j(t)$ capture the entire time dependence. 
Inserting this into the Heisenberg equation of motion and applying a sort of operator scalar product $(A_i|\cdot)$ 
on both sides of the equation yields
\bs
\begin{align}
        \ddt A(t) = \sum_n \ddt a_n(t) A_n &= \im \sum_n a_n(t) [H, A_n] \\
        \Rightarrow \sum_n \underbrace{(A_i | A_n)}_{\coloneqq \mN_{in}} \ddt a_n(t) &= 
				\im \sum_n \underbrace{(A_i | [H, A_n])}_{\coloneqq \mM_{in}} a_n(t) \\
        \Rightarrow \mN \ddt \vec{a}(t) &= \im \mM \vec{a}(t).
				    \label{eqn:heisenberg}
\end{align}
\es
where $\vec{a}:=(a_1, a_2, a_3,\ldots)^\top$.
The matrices $\mN$ and $\mM$ contain all the energetic and dynamic properties of the system.
The former is referred to as norm matrix while the latter is called the dynamic matrix.
The advantage is that one can now handle simple matrices rather than operators 
acting on an enormous Hilbert space.
The dimension of these matrices depends on the number of operators in the set $\mathcal{B}$.

@@

In order to treat the square and the simple cubic lattice on equal footing, we exploit 
\eqref{eqn:equal_expecs} to define the operator set in $\gamma$ space
\begin{equation}
    \label{eqn:ieom_basis_operator}
    A_\gamma \coloneqq \frac{1}{\sqrt{N}} \sum_{\vk} \delta (\gamma - \widehat{\gamma}( \vk )) A_{\vk},
\end{equation}
where $A_{\vk}$ represents each type of operator defined in Eqs.~\eqref{eqn:operators} or
their Hermitian conjugates if the operators are not Hermitian themselves.
Operators with different indices are treated separately, e.g., $n_{\gamma \up}$ and $n_{\gamma \down}$ 
are distinct operators in the studied set $\mathcal{B}$.
Additionally, we consider an analogous expression for the operator
\begin{equation}
    \tau_{k} \coloneqq c_{k \uparrow}^\dagger c_{k \downarrow}
\end{equation}
and its Hermitian conjugate.
This will allow us to describe transversal magnons, i.e., perpendicular 
deviations from the staggered magnetization in the AFM phase.

@@

In the AFM phase, the spectral functions $\spectral{l.AMF}$ and $\spectral{t.AFM}$ are distinct.
While the longitudinal magnon continues to manifest as a peak at relatively high energies
the transversal magnon appears as a peak at zero energy within our numerical accuracy. Due to
the asymmetry of the spectral, its peak is the derivative of a $\delta$ function.
It is an advantageous feature of the adopted approach that the Goldstone theorem is 
automatically fulfilled. Both, phase mode and transversal magnon are located at zero energy
if the corresponding symmetry is broken, i.e., the U(1) symmetry and the O(3) symmetry,
respectively.

@@

Additionally, we stress that expressions such as $(A_{\vk} | A_{\vl})$ or
$(A_{\vk} | [H, A_{\vl}])$ can be reduced to sums proportional to $\gamma_i$ or $\gamma_j$, respectively.
Therefore, it is not necessary to implement any sort of computation in reciprocal space.
As an example, let us consider $A_{\vk} = f_{\vk}^\dagger$ and 
\begin{equation}
    ( f_{\vk}^\dagger | f_{\vl}^\dagger ) = \delta_{\vk, \vl}\, \langle 1 - n_{\vk \uparrow} - n_{\vk \downarrow} \rangle = \alpha(\vk) \delta_{\vk, \vl}  
\end{equation}
with $\alpha(\vk)\coloneqq 1 - \langle n_{\vk \uparrow} + n_{\vk \downarrow} \rangle$. 
Inserting this into Eq.\ \eqref{eqn:numerical_example} yields
\begin{equation}
    (A_i | A_j) = \frac{1}{N} \sum_{\vk} \alpha(\vk) h(\gamma_i 
		- \widehat{\gamma}(\vk)) h(\gamma_j - \widehat{\gamma}(\vk)).
\end{equation}
As stated in the main text, see Eq.\ \eqref{eqn:equal_expecs}, $\alpha(\vk)$ depends only on $\widehat{\gamma}(\vk)$, 
hence we can pass from the wave vector integration to a $\gamma$ integration and eventually to a $\gamma$ sum
\bs
\begin{align}
    (A_i | A_j) &= \int \rho(\gamma) h(\gamma_i - \gamma) h(\gamma_j - \gamma) \alpha(\gamma) \dgamma 
      \\
    &\approx \rho(\gamma_i) h(\gamma_j - \gamma_i) \alpha(\gamma_i) 
			\\
    &= \frac{\delta_{ij}}{\Delta \gamma} \rho(\gamma_i) \alpha(\gamma_i).
\end{align}
\es
Since $\Delta \gamma$ is a positive constant factor, it does not impact any of our matrix operations, i.e., we can shift it to the front of the expressions $\mM \to \tilde{m} (\Delta \gamma)^2$ 
and $\mN \to \tilde{n} (\Delta \gamma)^2$.
This allows us to rewrite \eqref{eqn:green_derivation} as
\begin{equation}
    G_{AB^\dagger}(z) = - (\Delta \gamma)^2 \vec{b}^\dagger (0) \left[ \tilde{n} \frac{1}{\tilde{m} + z \tilde{n}} \tilde{n} \right] \vec{a}(0)
\end{equation}
for the general operators $A$ and $B$ and the initial conditions
$A=\sum a_i A_i$ and $B=\sum b_i A_i$.

@@

\bibliography{sn-bibliography}
%\bibliography{../../A_bibinput/liter10.bib}

>>>>

We start with an operator set $\mathcal{B}$ that ideally is complete with respect to commutation with 
the Hamiltonian $H$, i.e., any commutator $[H, A]$ for any $A \in \mathcal{B}$ can be represented by linear combinations of operators in $\mathcal{B}$.
Then, we can express any time-dependent operator in this set by
\begin{equation}
    \label{eqn:time_dependent_operator}
    A(t) = \sum_n a_n(t) A_n,
\end{equation}
where the coefficients $a_n(t)$ capture the entire time dependence. 
Inserting this into the Heisenberg equation of motion and applying a sort of operator scalar product $(A_i|\cdot)$ 
on both sides of the equation yields
\bs
\begin{align}
        \ddt A(t) = \sum_n \ddt a_n(t) A_n &= \im \sum_n a_n(t) [H, A_n] \\
        \Rightarrow \sum_n \underbrace{(A_i | A_n)}_{\coloneqq \mN_{in}} \ddt a_n(t) &= 
				\im \sum_n \underbrace{(A_i | [H, A_n])}_{\coloneqq \mM_{in}} a_n(t) \\
        \Rightarrow \mN \ddt \vec{a}(t) &= \im \mM \vec{a}(t).
				    \label{eqn:heisenberg}
\end{align}
\es
where $\vec{a}:=(a_1, a_2, a_3,\ldots)^\top$.
The matrices $\mN$ and $\mM$ contain all the energetic and dynamic properties of the system.
The former is referred to as norm matrix while the latter is called the dynamic matrix.
The advantage is that one can now handle simple matrices rather than operators 
acting on an enormous Hilbert space.
The dimension of these matrices depends on the number of operators in the set $\mathcal{B}$.

@@

In order to treat the square and the simple cubic lattice on equal footing, we exploit 
\eqref{eqn:equal_expecs} to define the operator set in $\gamma$ space
\begin{equation}
    \label{eqn:ieom_basis_operator}
    A_\gamma \coloneqq \frac{1}{\sqrt{N}} \sum_{\vk} \delta (\gamma - \widehat{\gamma}( \vk )) A_{\vk},
\end{equation}
where $A_{\vk}$ represents each type of operator defined in Eqs.~\eqref{eqn:operators} or
their Hermitian conjugates if the operators are not Hermitian themselves.
Operators with different indices are treated separately, e.g., $n_{\gamma \up}$ and $n_{\gamma \down}$ 
are distinct operators in the studied set $\mathcal{B}$.
Additionally, we consider an analogous expression for the operator
\begin{equation}
  \tau_k \coloneqq c_{\vec{k}\uparrow}^\dagger c_{\vec{k}+\vec{Q}\downarrow}
\end{equation}
and its Hermitian conjugate.
This will allow us to describe transversal magnons, i.e., perpendicular 
deviations from the staggered magnetization in the AFM phase.

@@

In the AFM phase, the spectral functions $\spectral{l.AFM}$ and $\spectral{t.AFM}$ are distinct.
While the longitudinal magnon continues to manifest as a peak at relatively high energies
the transversal magnon appears as a peak at zero energy within our numerical accuracy. Due to
the asymmetry of the spectral, its peak is the derivative of a $\delta$ function.
It is an advantageous feature of the adopted approach that the Goldstone theorem is 
automatically fulfilled. Both, phase mode and transversal magnon are located at zero energy
if the corresponding symmetry is broken, i.e., the U(1) symmetry and the O(3) symmetry,
respectively.

@@

Additionally, we stress that expressions such as $(A_{\vk} | A_{\vl})$ or
$(A_{\vk} | [H, A_{\vl}])$ can be reduced to sums proportional to $\gamma_i$ or $\gamma_j$, respectively.
Therefore, it is not necessary to implement any sort of computation in reciprocal space.
As an example, let us consider $A_{\vk} = f_{\vk}^\dagger$ and 
\begin{equation}
    ( f_{\vk}^\dagger | f_{\vl}^\dagger ) = \delta_{\vk, \vl}\, \langle 1 - n_{-\vk \uparrow} - n_{\vk \downarrow} \rangle = \alpha(\vk) \delta_{\vk, \vl}  
\end{equation}
with $\alpha(\vk)\coloneqq 1 - \langle n_{\vk \uparrow} + n_{-\vk \downarrow} \rangle$. 
Inserting this into Eq.\ \eqref{eqn:numerical_example} yields
\begin{equation}
    (A_i | A_j) = \frac{1}{N} \sum_{\vk} \alpha(\vk) h(\gamma_i 
		- \widehat{\gamma}(\vk)) h(\gamma_j - \widehat{\gamma}(\vk)).
\end{equation}
As stated in the main text, see Eq.\ \eqref{eqn:equal_expecs}, $\alpha(\vk)$ depends only on $\widehat{\gamma}(\vk)$, 
hence we can pass from the wave vector integration to a $\gamma$ integration and eventually to a $\gamma$ sum
\bs
\begin{align}
    (A_i | A_j) &= \int \rho(\gamma) h(\gamma_i - \gamma) h(\gamma_j - \gamma) \alpha(\gamma) \dgamma 
      \\
    &\approx \rho(\gamma_i) h(\gamma_j - \gamma_i) \alpha(\gamma_i) 
			\\
    &= \frac{\delta_{ij}}{\Delta \gamma} \rho(\gamma_i) \alpha(\gamma_i).
\end{align}
\es
Since $\Delta \gamma$ is a positive constant factor, it does not impact any of our matrix operations, i.e., we can shift it to the front of the expressions $\mM \to \tilde{m} (\Delta \gamma)^2$ 
and $\mN \to \tilde{n} (\Delta \gamma)^2$.
This allows us to rewrite \eqref{eqn:green_derivation} as
\begin{equation}
    G_{AB^\dagger}(z) = - (\Delta \gamma)^2 \vec{b}^\dagger (0) \left[ \tilde{n} \frac{1}{\tilde{m} + z \tilde{n}} \tilde{n} \right] \vec{a}(0)
\end{equation}
for the general operators $A$ and $B$ and the initial conditions
$A=\sum a_i A_i$ and $B=\sum b_i A_i$.

@@

\bibliography{main}
%\bibliography{../../A_bibinput/liter10.bib}
